\bf Model Visualization Techniques for a Social Network Model

### 2017-07-29 paper.Rnw
need to have most recent version of netvizif package to compile paper. also need dplyr version 0.5.0 for the unexported function match_data_frame

<<setup, echo = FALSE, message = FALSE, warning = FALSE>>=
options(replace.assign=TRUE,width=70, digits=3)
require(knitr)
opts_chunk$set(fig.path='figure/', cache.path='cache/', fig.align='center', fig.width=5, fig.height=5, par=TRUE, cache=TRUE, concordance=TRUE, autodep=TRUE, message=F, warning=F, echo = FALSE, root.dir = "~/Desktop/Dissertation - LETS GO/SAOM-removing-blindfold/")
@

@@

ggplot(data = toynet$edges) +
  geom_tile(aes(x = from, y = -to), color = 'grey20', fill = 'black') +
  #geom_tile(aes(x = to, y = from), color = 'grey20', fill = 'black') +
  geom_segment(aes(x = x, xend = xend), y = -0.5, yend=-5.5, colour="grey30", data = data.frame(x = 0:5+.5, xend= 0:5+.5)) + 
  geom_segment(aes(y=-y, yend=-yend), x = 0.5, xend=5.5, colour="grey30", data = data.frame(y = -0:5+.5, yend= -0:5+.5)) + 
  scale_x_continuous(breaks = 1:5, labels = 1:5, limits = c(0.5,5.5), position = "top") +
  scale_y_continuous(breaks = (-5):(-1), labels = 5:1, limits = c(-5.5,-.5)) +
  theme_bw() + 
  theme(panel.grid = element_blank()) +
  labs(title = "Adjacency Matrix Visualization") + 
  theme(aspect.ratio = 1) +
  xlab("from (ego)") +
  ylab("to (alter)")

@
\hh{XXX Explain each of these visualizations and either comment on the differences in the visualizations (colors) or don't show them. }

@@

Using the distribution given in Equation~\ref{dist1} with mean $\mu = 0$ and scale $\sigma = 1$ as \citet{modelsSnijders} suggests \hh{XXX I'm not following - which distribution do you mean by `this' distribution? The density above is the type 1 extreme value density. I got lost at the  ``objective function $f_i(\boldsymbol{\beta}, x)$ plus a random element, $U_i(x)$"} \st{XXX I hope it is clearer now above}
is convenient because its \st{properties lead to a probability formula for} the probability that actor $i$ chooses to change its tie to actor $j$ \st{that can be written only} in terms of the objective function. Let $p_{ij}(\boldsymbol{\beta}, x)$ be the probability that actor $i$ chooses to change its tie to actor $j$. Next, we write the network $x$ in its potential future state, where the tie $x_{ij}$ has changed to $1-x_{ij}$, as $x(i \leadsto j)$. Then, the probility that the tie $x_{ij}$ changes is

@@

# running Siena
#friendship <- sienaDependent(friendshipData)
#alcohol <- varCovar(drink[20:35,])
#mydata <- sienaDataCreate(friendship, alcohol)
#myeffnull <- getEffects(mydata)
#myalgorithm <- sienaAlgorithmCreate(projname = 's16_3')
set.seed(823746)
ansnullchains <- get_chain_info(ansnull)
ansnullchains %>%
  filter(period == 1) %>%  #only look at chains from wave 1 to wave 2
  group_by(rep) %>%
  dplyr::select(rep, from = ego, to = alter) %>%
  mutate(val = as.numeric(!from == to),
         from = paste0("V", parse_number(from)+1), # make the chains
         to = paste0("V", parse_number(to)+1)) -> ansnullchainsw1w2
colnames(fd2.w1) <- paste0("V", 1:16)
rownames(fd2.w1) <- paste0("V", 1:16)
wave1friends <- fortify(as.adjmat(fd2.w1))
ms1 <- listMicrosteps(dat = wave1friends,
                      microsteps = filter(ansnullchainsw1w2, rep == 1))
microsteps <- ms1[1:20]

@@

<<histogramalters, fig.width = 8, fig.height=11, fig.cap="Each panel shows the theoretical (as bars) and empirical (as points) probabilities of the chosen ego node changing its tie to each of the other nodes.">>=
ansnullchains %>% group_by(period, rep) %>% 
  mutate(ms = row_number()) %>% 
  filter(ms == 1, period == 1) %>% 
  mutate(alterProb = exp(logAlterProb),
         ego = as.numeric(as.character(ego)),
         alter = as.numeric(as.character(alter))) %>% 
  ungroup() %>% 
  group_by(ego, alter) %>% 
  summarize(count = n(), alterProb = unique(alterProb)) %>% ungroup() %>% 
  group_by(ego) %>% 
  mutate(csum = cumsum(alterProb)) %>% 
  arrange(ego, csum) %>% 
  mutate(totego = sum(count), 
         empprob = count / totego) -> cumprobs 
cumprobs$ego <- factor(paste0("V",cumprobs$ego+1), levels = paste0("V", 1:16))
cumprobs$alter <- factor(paste0("V",cumprobs$alter+1), levels = paste0("V", 1:16))

@@

cumprobs %>% left_join(changes2) %>% ggplot() + 
  geom_histogram(
    aes(x = alter, weight = alterProb, fill = type), 
    color = 'black', fill = 'white', binwidth = 1) + 
  geom_point(aes(x = alter, y = empprob), color = 'red') + 
  scale_x_continuous(labels = paste0("V",1:16), breaks = 0:15) + coord_flip() + facet_wrap(~ego, labeller = "label_both") + theme_gray()
# 
# cumprobs %>% ggplot() + geom_histogram(aes(x = alter, weight = count), color = 'black', fill = 'white', binwidth = 1) + scale_x_continuous(labels = paste0("V",1:16), breaks = 0:15) + coord_flip() + facet_wrap(~ego, labeller = "label_both") + theme_gray()

@@

Another way to view the transition probabilities is through the adjacency matrix visualization. In Figure~\ref{fig:heatmap}, we build on the concept of the ordered adjacency matrix of Figure~\ref{fig:adjmatorder}. This heatmap shows the transition probabilities of all ties that are changed in the first microstep of the 1,000 simulations. The heatmap is noticeably sparse, showing the lack of coverage in the model.  Of the 256 possible steps, only 81, or about 32\%, are taken in the 1,000 simulated chains. XXX Check this again.... XXX
<<heatmap>>=
ansnullchains %>% 
  filter(rep == 1, period == 1) %>% 
  mutate(ego = paste0("V",as.numeric(as.character(ego))+1),
         alter = paste0("V",as.numeric(as.character(alter))+1), 
         prob = exp(logAlterProb)) %>% 
  ggplot(aes(x = factor(ego, levels = order1), 
             y = factor(alter, levels = rev(order1)), 
             fill = prob)) + 
  geom_tile(color = 'black') + 
  scale_fill_gradient(low = 'white', high = 'blue', limits = c(0,1)) +
  theme(panel.background = element_rect(fill = 'white')) + 
  labs(x = "ego", y = "alter")+
  coord_fixed()
@

@@

<<summnetM1>>=
friendshipData <- array(c(fd2.w1, fd2.w2, fd2.w3), dim = c(16, 16, 3))
drink1 <- s50a[20:35,1]
drink2 <- s50a[20:35,2]
drink3 <- s50a[20:35,3]
friendship <- sienaDependent(friendshipData)
alcohol <- varCovar(cbind(drink1,drink2, drink3))
mydata <- sienaDataCreate(friendship, alcohol)
M1eff <- getEffects(mydata)
M2eff <- includeEffects(M1eff, jumpXTransTrip, interaction1 = "alcohol")
M3eff <- includeEffects(M1eff, nbrDist2twice)
means <- simu2 %>% group_by(Model, parameter) %>% summarize(
  means = mean(estimate)
)
M1parms <- (means %>% filter(Model == "M1"))$means
M2parms <- (means %>% filter(Model == "M2"))$means
M3parms <- (means %>% filter(Model == "M3"))$means
set.seed(4231352)
M1sims1000 <- saom_simulate(dat = mydata, 
                            struct = M1eff,
                       parms = as.numeric(M1parms), 
                       N = 1000)
M2sims1000 <- saom_simulate(dat = mydata, 
                            struct = M2eff,
                       parms = as.numeric(M2parms), 
                       N = 1000)
M3sims1000 <- saom_simulate(dat = mydata, 
                            struct = M3eff,
                       parms = as.numeric(M3parms), 
                       N = 1000)
M1simsdf <- sims_to_df(M1sims1000)
M2simsdf <- sims_to_df(M2sims1000)
M3simsdf <- sims_to_df(M3sims1000)
M1avgW2 <- M1simsdf %>% filter(!is.na(to) & wave == 2) %>%
  group_by(from, to) %>% 
  summarise(count = n()) %>% 
  mutate(weight = ifelse(from == to, 0, count))
M2avgW2 <- M2simsdf %>% filter(!is.na(to) & wave == 2) %>%
  group_by(from, to) %>% 
  summarise(count = n()) %>% 
  mutate(weight = ifelse(from == to, 0, count))
M3avgW2 <- M3simsdf %>% filter(!is.na(to) & wave == 2) %>%
  group_by(from, to) %>% 
  summarise(count = n()) %>% 
  mutate(weight = ifelse(from == to, 0, count))

@@

simu2 <- read.csv("data/simulation-1000-M1-M2-M3.csv")

>>>>

<<setup, echo = FALSE, message = FALSE, warning = FALSE>>=
options(replace.assign=TRUE,width=70, digits=3)
require(knitr)
opts_chunk$set(fig.path='figure/', cache.path='cache/', fig.align='center', fig.pos='h', fig.width=5, fig.height=5, par=TRUE, cache=TRUE, concordance=TRUE, autodep=TRUE, message=F, warning=F, echo = FALSE, root.dir = "~/Desktop/Dissertation - LETS GO/SAOM-removing-blindfold/")
@

@@

ggplot(data = toynet$edges) +
  geom_tile(aes(x = from, y = -to), color = 'grey20', fill = 'black') +
  #geom_tile(aes(x = to, y = from), color = 'grey20', fill = 'black') +
  geom_segment(aes(x = x, xend = xend), y = -0.5, yend=-5.5, colour="grey30", data = data.frame(x = 0:5+.5, xend= 0:5+.5)) + 
  geom_segment(aes(y=-y, yend=-yend), x = 0.5, xend=5.5, colour="grey30", data = data.frame(y = -0:5+.5, yend= -0:5+.5)) + 
  scale_x_continuous(breaks = 1:5, labels = 1:5, limits = c(0.5,5.5), position = "top") +
  scale_y_continuous(breaks = (-5):(-1), labels = 5:1, limits = c(-5.5,-.5)) +
  theme_bw() + 
  theme(panel.grid = element_blank()) +
  labs(title = "Adjacency Matrix Visualization") + 
  theme(aspect.ratio = 1) +
  xlab("from (ego)") +
  ylab("to (alter)")
@
\hh{XXX Explain each of these visualizations and either comment on the differences in the visualizations (colors) or don't show them. }

@@

Using the distribution given in Equation~\ref{eq:dist1} with mean $\mu = 0$ and scale $\sigma = 1$ as \citet{modelsSnijders} suggests \hh{XXX I'm not following - which distribution do you mean by `this' distribution? The density above is the type 1 extreme value density. I got lost at the  ``objective function $f_i(\boldsymbol{\beta}, x)$ plus a random element, $U_i(x)$"} \st{XXX I hope it is clearer now above}
is convenient because its \st{properties lead to a probability formula for} the probability that actor $i$ chooses to change its tie to actor $j$ \st{that can be written only} in terms of the objective function. Let $p_{ij}(\boldsymbol{\beta}, x)$ be the probability that actor $i$ chooses to change its tie to actor $j$. Next, we write the network $x$ in its potential future state, where the tie $x_{ij}$ has changed to $1-x_{ij}$, as $x(i \leadsto j)$. Then, the probility that the tie $x_{ij}$ changes is

@@

# running Siena
#friendship <- sienaDependent(friendshipData)
#alcohol <- varCovar(drink[20:35,])
#mydata <- sienaDataCreate(friendship, alcohol)
#myeffnull <- getEffects(mydata)
#myalgorithm <- sienaAlgorithmCreate(projname = 's16_3')
set.seed(823746)
.libPaths(new = c("/Library/Frameworks/R.framework/Versions/3.4/Resources/library", "/Library/Frameworks/R.framework/Versions/3.4/Resources/oldpkgs/"))
ansnullchains <- get_chain_info(ansnull)
ansnullchains %>%
  filter(period == 1) %>%  #only look at chains from wave 1 to wave 2
  group_by(rep) %>%
  dplyr::select(rep, from = ego, to = alter) %>%
  mutate(val = as.numeric(!from == to),
         from = paste0("V", parse_number(from)+1), # make the chains
         to = paste0("V", parse_number(to)+1)) -> ansnullchainsw1w2
colnames(fd2.w1) <- paste0("V", 1:16)
rownames(fd2.w1) <- paste0("V", 1:16)
wave1friends <- fortify(as.adjmat(fd2.w1))
ms1 <- listMicrosteps(dat = wave1friends,
                      microsteps = filter(ansnullchainsw1w2, rep == 1))
microsteps <- ms1[1:20]

@@

<<histogramalters, fig.width = 8, fig.height=11, fig.cap="Each panel shows the theoretical (as lines) and empirical (as points) probabilities of the chosen ego node changing its tie to each of the other nodes. The color of the line indicates whether the tie from the ego to the alter node is being added, removed, or if there is no change to the network in this step.">>=
ansnullchains %>% group_by(period, rep) %>% 
  mutate(ms = row_number()) %>% 
  filter(ms == 1, period == 1) %>% 
  mutate(alterProb = exp(logAlterProb),
         ego = as.numeric(as.character(ego)),
         alter = as.numeric(as.character(alter))) %>% 
  ungroup() %>% 
  group_by(ego, alter) %>% 
  summarize(count = n(), alterProb = unique(alterProb)) %>% ungroup() %>% 
  group_by(ego) %>% 
  mutate(csum = cumsum(alterProb)) %>% 
  arrange(ego, csum) %>% 
  mutate(totego = sum(count), 
         empprob = count / totego) -> cumprobs 
cumprobs$ego <- factor(paste0("V",cumprobs$ego+1), levels = paste0("V", 1:16))
cumprobs$alter <- factor(paste0("V",cumprobs$alter+1), levels = paste0("V", 1:16))

@@

cumprobs %>% left_join(changes2) -> changes3

changes3$ego <- factor(changes3$ego, levels = paste0("V", 1:16), labels = 1:16) 
changes3$alter <- factor(changes3$alter, levels = paste0("V", 1:16), labels = 1:16)

ggplot(data = changes3) + 
  geom_segment(
    aes(x = alter , xend = alter, y = 0, yend = alterProb, color=type)) + 
  geom_point(aes(x = alter, y = empprob), shape = 4, size = 1, alpha = .5) + 
  labs(x = "alter node", y = "probability of alter node") + 
  coord_flip() + 
  facet_wrap(~ego, labeller = "label_both") + 
  theme_bw()
# 
# cumprobs %>% ggplot() + geom_histogram(aes(x = alter, weight = count), color = 'black', fill = 'white', binwidth = 1) + scale_x_continuous(labels = paste0("V",1:16), breaks = 0:15) + coord_flip() + facet_wrap(~ego, labeller = "label_both") + theme_gray()

@@

Another way to view the transition probabilities is through the adjacency matrix visualization. In Figure~\ref{fig:heatmap}, we build on the concept of the ordered adjacency matrix of Figure~\ref{fig:adjmatorder}. This heatmap shows the transition probabilities of all ties that are changed in the first microstep of the 1,000 simulations. The heatmap is noticeably sparse, showing the lack of coverage in the model.  Of the 256 possible steps, only 81, or about 32\%, are taken in the 1,000 simulated chains. XXX Check this again.... XXX
<<heatmap, fig.align='center', fig.cap='A heatmap showing the empirical transition probabilities for the first microstep in 1,000 simulations. The acting node is on the x-axis, and the receiving node is on the y-axis. Missing squares represent ties that did not occur.'>>=
ansnullchains %>% 
  filter(rep == 1, period == 1) %>% 
  mutate(ego = paste0("V",as.numeric(as.character(ego))+1),
         alter = paste0("V",as.numeric(as.character(alter))+1), 
         prob = exp(logAlterProb)) -> probsDat

ggplot(data = probsDat, 
       aes(x = factor(ego, levels = order1), 
             y = factor(alter, levels = rev(order1)), 
             fill = prob)) + 
  geom_tile(color = 'black') + 
  scale_fill_gradient(low = 'white', high = 'blue', limits = c(0,1)) +
  theme(panel.background = element_rect(fill = 'white')) + 
  labs(x = "ego", y = "alter")+
  scale_x_discrete(position = 'top') + 
  coord_fixed()
@

@@

% below chunk contains some repetitive stuff. Haven't figured out what it's purpose is yet. 
<<summnetM1, message=FALSE>>=
friendshipData <- array(c(fd2.w1, fd2.w2, fd2.w3), dim = c(16, 16, 3))
drink1 <- s50a[20:35,1]
drink2 <- s50a[20:35,2]
drink3 <- s50a[20:35,3]
friendship <- sienaDependent(friendshipData)
alcohol <- varCovar(cbind(drink1,drink2, drink3))
mydata <- sienaDataCreate(friendship, alcohol)
M1eff <- getEffects(mydata)
M2eff <- includeEffects(M1eff, jumpXTransTrip, interaction1 = "alcohol")
M3eff <- includeEffects(M1eff, nbrDist2twice)
simu2 <- read.csv("data/simulation-1000-M1-M2-M3.csv")
means <- simu2 %>% group_by(Model, parameter) %>% summarize(
  means = mean(estimate)
)
M1parms <- (means %>% filter(Model == "M1"))$means
M2parms <- (means %>% filter(Model == "M2"))$means
M3parms <- (means %>% filter(Model == "M3"))$means
set.seed(4231352)
M1sims1000 <- netvizinf::saom_simulate(dat = mydata, 
                            struct = M1eff,
                       parms = as.numeric(M1parms), 
                       N = 1000)
M2sims1000 <- netvizinf::saom_simulate(dat = mydata, 
                            struct = M2eff,
                       parms = as.numeric(M2parms), 
                       N = 1000)
M3sims1000 <- netvizinf::saom_simulate(dat = mydata, 
                            struct = M3eff,
                       parms = as.numeric(M3parms), 
                       N = 1000)
M1simsdf <- sims_to_df(M1sims1000)
M2simsdf <- sims_to_df(M2sims1000)
M3simsdf <- sims_to_df(M3sims1000)
M1avgW2 <- M1simsdf %>% filter(!is.na(to) & wave == 2) %>%
  group_by(from, to) %>% 
  summarise(count = n()) %>% 
  mutate(weight = ifelse(from == to, 0, count))
M2avgW2 <- M2simsdf %>% filter(!is.na(to) & wave == 2) %>%
  group_by(from, to) %>% 
  summarise(count = n()) %>% 
  mutate(weight = ifelse(from == to, 0, count))
M3avgW2 <- M3simsdf %>% filter(!is.na(to) & wave == 2) %>%
  group_by(from, to) %>% 
  summarise(count = n()) %>% 
  mutate(weight = ifelse(from == to, 0, count))

@@
